---
title: "cm103 Worksheet"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse)) # Loads purrr, too
library(repurrrsive) # Contains data examples
library(listviewer) # For viewing JSON/lists interactively
```

## Resources

This week, we'll be drawing from [Jenny Bryan's `purrr` tutorial](https://jennybc.github.io/purrr-tutorial/). Specifically:

- The [`map` tutorial](https://jennybc.github.io/purrr-tutorial/ls01_map-name-position-shortcuts.html)
- The [GitHub users tutorial](https://jennybc.github.io/purrr-tutorial/ls02_map-extraction-advanced.html) is similar.

In addition:

- Do you feel that you need a refresher on lists and vectors in R? Check out the [Vectors and Lists](https://jennybc.github.io/purrr-tutorial/bk00_vectors-and-lists.html) part of the tutorial.
- Are you familiar with the `apply` family of functions in Base R? You might find Jenny's ["Relationship to base and plyr functions"](https://jennybc.github.io/purrr-tutorial/bk01_base-functions.html) great for bridging the gap to `purrr`.


## Review Vectors and Lists

Vectors: 

- hold multiple entries of the same type.
- coerced to the least-informative data type in the vector.
- subset with single square brackets

Lists:

- hold multiple entries of anything.
- no entries are coerced (as a result of being able to hold anything)
- subset with `[`, `[[`, or `$`. 
single [ will give you the first car], and [[ will give you the content of the first car ]]

## Review Vectorization

= element-wise application of a function.

Examples:

```{r}
(1:10) ^ 2
```

```{r}
(1:10) * (1:2)
```

```{r}
commute <- c(40, 20, 35, 15)
person <- c("Parveen", "Leo", "Shawn", "Emmotions")
str_c(person, " takes ", commute, " minutes to commute to work.")
```

## `purrr`

`purrr` is great when vectorization does not apply! 

Particularly useful if your data is in JSON format. 

Example: 

1. Explore the `wesanderson` list (comes with `repurrrsive` package). Hint: `str()` might help. It's a list of vectors. 
2. Use what you know about R to write code that extracts a vector of the first elements of each vector contained in `wesanderson`.

```{r}


#this returns the first component in the first vector.
wesanderson[[1]][1]

#use a for loop:
x = character(0)
#character vector of length 0

for (i in 1:length(wesanderson)) x[i] = wesanderson[[i]][1]

x



```

`str()` is not always useful! Try checking the structure of `got_chars` (= Game of Thrones characters):

```{r}
#first top level entry is a cahracter of GOT. each character has its own list of 18 too. 
str(got_chars) 
#max levels only gives you the first component of the list. max level 2 gives the sublists.
str(got_chars, list.len = 4, max.level = 1)
str(got_chars, list.len = 4, max.level = 2)



```


## Exploring lists

1. `str()`: embrace `list.len` and `max.level`

```{r}
str(got_chars)
```

2. Interactive exploration: `View()` and `listviewer::jsonedit(..., mode = "view")`

```{r}
View(got_chars)
jsonedit(got_chars, mode="view")


```

3. Don't be afraid to check out a subset! `names()` comes in handy, too.

```{r}
str(got_chars[[1]])
#what information does the first person have about them.
names(got_chars[[1]])
```


## Exploring `purrr` fundamentals

Apply a function to each element in a list/vector with `map`.

General usage: `purrr::map(VECTOR_OR_LIST, YOUR_FUNCTION)`

Note:

- `map` always returns a list.
- `YOUR_FUNCTION` can return anything!

Toy example 1: without using vectorization, take the square root of the following vector:

```{r}
x <- 1:10
#if it's vectorized we can do the following to take sqrt of each element: 

sqrt(x)

#if it's not vectorized, then we could use map to do the same thing.
map(x, sqrt)
```

Toy example 2 (functions on-the-fly): without using vectorization, square each component of `x`:

```{r}
square = function(x) x^2
map(x, square)

#we can also define our own function within the brackets:
map(x, function(w) w^2)
```

Want a vector to be returned? Must specify the `typeof()` of the vector. Use `map_dbl()` to specify an output vector of type "double" for the above (check out the documentation `?map` for the acceptable vector types):

```{r}
#what type of vector output do we want? use map_* where * = something like dbl, chr, lgl = logical? , integer... etc.
map_dbl(x, sqrt)
map_chr(x, sqrt)
```

Does your function have other arguments? You can specify them afterwards in the ellipses (`...`).

```{r}
#can specify more arguments.
map_chr(x, str_c, "potato.", sep="-")
```

## Your Turn: `purrr` fundamentals

1. Let's retry the `wesanderson` example: use `purrr` to write code that extracts a vector of the first elements of each vector contained in `wesanderson`. Play around with the following, too:
    - Use `head` instead of writing your own function.
    - Try different `map` functions, even the "wrong" types.
    - Use the ` ``[`` ` function if you're feeling daring.

```{r}
first_element = function(x) x[[1]][1]
map(wesanderson, first_element)

#class solution:
#use map_chr because we want a character output.
vec1 = function(x) x[1] #takes the vectors first element
map_chr(wesanderson, vec1)

#class solution 2: head
map_chr(wesanderson, head, n = 1)

#class solution 3: 
#when using `[` we use the format: `[`(x,1) where x is what you are subsetting and 1 is that you want the first element of it. here we only use 1 and no x, since we already specify wesanderson in map_chr call.
map_chr(wesanderson, `[`,1)

#backtick marks work for other functions too:
y = function(x)x^2
`y`(5) #returns 25


```

2. Check that each character's list entry in `got_chars` has the same names as everyone else (that is, list component names, not character names). Here's one way to do it:
    1. Use the `names` function.
    2. Then, bind the names together in a single character.
    3. Then, apply the `table()` function.

```{r}
#skip this exercise.
```


## Shortcut functions

We can do the subsetting much easier with these shortcuts: just replace function with either:

- index you'd like to subset by, or
- name you'd like to subset by.

```{r}
#just putting an integer automatically knows to take first component of each var.
map_chr(wesanderson, 1) # %>% unname()

#if don't want names to be there, use unname()
map_chr(wesanderson, 1) %>% 
  unname()
```


## Your turn: shortcut functions

1. For the `got_chars` data:

- What are the titles of each character? 
- Is a vector output appropriate here?
- Use a pipe.

Note: each character's list entry has a component named `titles` as the 9th entry.

```{r}
#use map bc each character can have more than 1 title. vector output is not appropriate bc we don't know how long each output will be

got_chars %>% 
  map("titles") #map(9) would return the same thing.

```

2. For the `got_chars` data:

- Extract a list of the "name" and "born" data for each person.
    - Use the function ` ``[`` ` or `extract()` (from the `magrittr` package, does the same thing) function to do the subsetting
- What happens when we switch to `map_dfr` instead of `map`? How about `map_dfc`?

```{r}
desired_info <- c("name", "born")
#use subset `[` to subset by desired info
map(got_chars, `[`, desired_info) #keeps same structure as original got_chars data but removed list components for each person. 

map(got_chars, `[`, desired_info) %>% 
  head()

#convert json format data into a df: each person now has their own row, and info from each person now found in each column.
#not best idea to use this map_dfr all the time... it's more error prone.

map_dfr(got_chars, `[`, desired_info)
```

Note: as Jenny says, it's always safer (from a programming perspective) to work with output that's more predictable. The following would be safer, and is still readable:

```{r}
#instead of using map_dfr, can use piping to send to tibble function:

got_chars %>% { #use of {} to 'trick' R because with PIPE it pipes into the first argument of the function. 
    tibble(
      #the . means whatever is on LHS of the pipe into the first function. then we subset to name and convert that to a character vector of names. 
        name = map_chr(., "name"),
        
        born = map_chr(., "born")
    )
}

#if didn't use {} it would give a 3 column dataframe, where first column returns a list. why? got_chars gets piped into first argument of tibble, which is its own column. then it thinks we want to make more columns on top of that first one.

?`{`
#curly braces -> can have multiple lines. evals and returns only the last thing that was evaluated within the curly braces. so it's as if we pipe got_chars into the first function call, which is either the curly braces or tibble depeneding on whether we use { } or not.
```

Note the curly braces "tricks" the object prior to the pipe from entering as the first argument to `tibble`, because ` ``{`` ` only returns the last line evaluated. In this case there are two: the above code is equivalent to

```{r}
got_chars %>% {
    .
    tibble(
        name = map_chr(., "name"),
        born = map_chr(., "born")
    )
}
```

