---
title: "cm008 Exercises"
output: github_document
---

## Install `nycflights13` package

```{r,eval=FALSE}
install.packages("nycflights13")
```

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(nycflights13))
```

## Types of mutating join

### Let's join tibbles using four mutating functions:  `left_join`, `right_join`, `inner_join` and `full_join`.

### create two tibbles named `a` and `b`

```{r}
(a <- tibble(x1 = LETTERS[1:3], x2 = 1:3))
(b <- tibble(x1 = LETTERS[c(1,2,4)], x3 = c("T", "F", "T")))
```

### left_join: Join matching rows from `b` to `a` by matching "x1" variable

```{r}
#prints variables that match to x1.
left_join(a,b,by="x1")

```


### right_join: Join matching rows from `a` to `b` by matching "x1" variable.

```{r}
#prints variables that match 
right_join(a,b,by="x1")
```



### inner_join: Join data. Retain only rows in both sets `a` to `b` by matching "x1" variable.

```{r}
inner_join(a, b, by = "x1")
```


### full_join: Join data. Retain all values, all rows of `a` to `b` by matching "x1"

```{r}
full_join(a, b, by = "x1")
```


### what happen if we do not specify `by` option?

```{r}
left_join(a,b)
#mutates by matching columns (by column name), as x1 is found in both 
#cannot combine more than 2 tibbles, but reduce package takes all the tibbles and does this function in a loop.
```


### what happen if we join two different variables (e.g., "x1" to "x3") from two tibbles `a` to `b`?
```{r}
left_join(a,b,by= c("x1" = "x3")) #no match, so it prints only NA. 
b
a
#we can specify which column to take from left and which from right.  (see below?)
```



### what happen if two columns of `a` and `c` datasets have the identical colnames?

```{r}

# make data frame c and use inner_join()
(c <- tibble(x1 = c(LETTERS[1:2],"x"), x2 = c(1,4,5)))
left_join(a,c)
#how does dplyr handle identical column names?

```

## In class practice

`nycflights13` dataset has several tibbles e.g., `flights`, `airports`, `planes`, `weather`.

### 1. Explore `nycflights13` dataset

```{r}
#check the tibbles included in `nycflights13` package
class(flights)

colnames(flights) #19 columns in flights

colnames(airlines) #two columns

colnames(weather)
```

### 2. Drop unimportant variables so it's easier to understand the join results. Also take first 1000 rows to run it faster.

```{r}
flights2 <- flights[1:1000,] %>% 
  select(year, tailnum, carrier, time_hour)
dim(flights2)
```

### 3. Add airline names to `flights2` from `airlines` dataset.

```{r}
# Which join function to use?
airlines
flights2

left_join(flights2, airlines) #airline names are added from the airline dataset.  using left join we added airline names joined by carrier column. no need to specify as it automatically joins by common variable???

```

### 4. Add `weather` information to the `flights2` dataset by matching "year" and "time_hour" variables.

```{r}
weather
left_join(flights2, weather, by = c("year", "time_hour"))
#why does c() work in this case but in case of filtering data it did not work? 
#another way to do this
flights2 %>% 
  left_join(weather, by=c("year", "time_hour"))
```


### 5. Add `weather` information to the `flights2` dataset by matching only "time_hour" variable

```{r}
flights %>% 
  left_join(weather, by = "time_hour")
```


## Types of filtering join

### Let's filter tibbles using two filtering functions: `semi_join`, `anti_join` 

```{r}
#semi join = filter (keep) all rows in a that have a match in b. 
semi_join(a,b)
#anti-join = filter (keep) all rows in a that do not have a match in b.
anti_join(a,b)
#we can also specify which variable we want to filter by:
anti_join(a,b, by="x1") #but in thiscase it's automatcially done this way bc x1 is the only matching 
```


### example for `semi_join`: All rows in `a` that have a match in `b`

see above.

### example for `anti_join`: All rows in `a` that do not have a match in `b`

see above.

### example of joinin by matching two variables (e.g., "x1", "x2") from both datasets `a` and `c`

```{r}
a
c
#a and c have identical column names
semi_join(a,c) #so automatically it will filter both by x1 and x2, therefore keeping only row A, as this is the only one that matched in both variables x1 and x2. 
```


## Types of Set Operations for two datasets

### Let's use three `set` functions: `intersect`, `union` and `setdiff` 

### create two tibbles named `y` and `z`, similar to Data Wrangling Cheatsheet

```{r}
(y <-  tibble(x1 = LETTERS[1:3], x2 = 1:3))
(z <- tibble(x1 = c("B", "C", "D"), x2 = 2:4))
```

### example for `intersect`: Rows that appear in both `y` and `z`

```{r}
intersect(y,z)
```


### example for `union`: Rows that appear in either or both `y` and `z`

```{r}
union(y,z)
```


### example for `setdiff`: Rows that appear in `y` but not `z`. __Caution:__ `setdiff` for `y` to `z` and `z` to `y` are different.


```{r}
setdiff(y,z) #rows that appear in y but not z
setdiff(z,y) #rows that appear in z but not y.
```

### what happen if colnames are differentin `y` and `x`? Is there any error message and why?

```{r}
(x <- tibble(x1 = c("B", "C", "D"), x3 = 2:4))
#intersect(y,x) if we run this we have an error, because there neeeds to be identical columns. this issn't the case here. 
#union(x,y)

```

## Types of binding datasets

### Let's bind datasets by rows or column using two binding functions:

### example for `bind_rows`: Append `z` to `y` as new rows

```{r}
bind_rows(y,z) #adds all the rows of z to bottom of y.
```


### example for `bind_cols`: Append `z` to `y` as new columns. __Caution__: matches rows by position. Check colnames after binding.

```{r}
bind_cols(y,z) #notice the change of column names.
```


### what happen if colnames are different between `y` and `x` datasets?

```{r}
bind_cols(y,x) #there is no conflict for column binding. no need for the renaming of column name. 
bind_rows(y,x)
```

##Practice Exercises
Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.


### 1. Filter the rows of `flights2` by matching "year" and "time_hour" variables to `weather` dataset. Use both `semi_join()` and `anti_join()`

```{r}
semi_join(flights2, weather, by=c("year", "time_hour"))
anti_join(flights2, weather, by=c("year", "time_hour")) #see no mismatch. 
```


### 2. Can we apply `set` and `binding` funcions between `flights2` and `weather` datasets. Why and why not?
```{r}
union(flights2, weather) #error bc col names in flight 2 and weather are different. 
intersect(flights2, weather)
setdiff(flights2, weather)
bind_cols(flights2, weather) #doesn't work bc diff number of rows.
bind_rows(flights2, weather) #this should work? 
#none of them work? why?? 
```


### 3. Let's create a tibble `p` with "x1" and "x2" coulmns and have duplicated element in "x1" column. Create another tibble `q` with "x1" and "x3" columns. Then apply `left_join` function `p` to `q` and `q` to `p`. 

```{r}
#(p <- tibble(x1 = LETTERS[1:3], x2 = LETTERS[1:3]))
#q <- tibble(x1 = LETTERS[1:3], x3 = 1:3))
#left_join(p,q)
#left_join(q,p)

(p = tibble(x1 = c("a", "b", "b", "c"),x2 = 1:4))
(q = tibble(x1 = c("a", "b", "c", "c"), x3= 5:8))

left_join(p,q) #b is found twice in dataset p, so it prints the value twice in q.
left_join(q,p) #c is found twice in q so when joining it prints the value 4 in p twice.  


```

