---
title: "cm006 Exercises: Exploring Geometric Objects"
output: github_document
editor_options: 
  chunk_output_type: inline
---

In this worksheet, we'll be exploring various plot types (i.e., geometric objects), only using the `x` and `y` aesthetics (and `group`).

1. To get started, load the `tidyverse` and `gapminder` R packages. 

```{r}
library(tidyverse) #load tidyverse in which ggplot2 is found
library(gapminder) #load gapminder data
head(gapminder
     )

```

## Scatterplot

Let's look at a _scatterplot_ of `gdpPercap` vs. `lifeExp`. 

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.
    - We'll ignore "coordinate system" and "facetting" after this.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ | 'x' and 'y'|
| __geometric object__  | point |
| scale                 | linear (default) |
| statistical transform | no transformation|
| coordinate system     | cartesian // rectangular |
| facetting             |  none |

2. Populate the data and aesthetic mapping in `ggplot`. What is returned? What's missing?

```{r}
ggplot(gapminder, aes(x=lifeExp, y=gdpPercap)) + geom_point()
```


3. Add the missing component as a _layer_.

--> The added layer is the +geom_point() and was added in part 2. 

Notice the "metaprogramming" again!

--> no need to use ' ' when calling the lifeExp and gdpPercap. this is due to metaprogramming.

4. You _must_ remember to put the aesthetic mappings in the `aes` function! What happens if you forget?

```{r}
#ggplot(gapminder, x=lifeExp, y=gdpPercap) + geom_point()
#this will give an error. need aes function with minimal x and y data. can also change colour etc here, but not needed.

ggplot(gapminder, aes(x=lifeExp, y=gdpPercap)) + geom_point()

```


5. Put the x-axis on a log scale, first by transforming the x variable. 
    - Note: `ggplot2` does some data wrangling and computations itself! We don't always have to modify the data frame.

```{r}
ggplot(gapminder, aes(x=lifeExp, y=log(gdpPercap))) + geom_point() #first arg for aes are automatically x and y unless you specify. x= y=

```


6. Try again, this time by changing the _scale_ (this way is better).

```{r}
ggplot(gapminder, aes(x=lifeExp, y=gdpPercap)) +
  geom_point() +
  scale_y_log10()

```


7. The aesthetic mappings can be specified on the geom layer if you want, instead of the main `ggplot` call. Give it a try:


8. Optional: git stage and commit

__Uses of a scatterplot__: 

- Visualize 2-dimensional distributions; dependence.
- 2 numeric variables

## Histograms, and Kernel Density Plots

Let's build a histogram of life expectancy.

-> useful to visualize distribution of a numeric variable. 

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ |  'x'|
| __geometric object__  |  histogram |
| scale                 |  linear|
| statistical transform |  none |

2. Build the histogram of life expectancy.

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram()
```


3. Change the number of bins to 50.

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(bins = 50)
  
  
```


4. Instead of a histogram, let's create a kernel density plot.

```{r}
ggplot(gapminder, aes(lifeExp)) + geom_density()

#can change the bin width by changing bw. 
ggplot(gapminder, aes(lifeExp)) + geom_density(bw = 0.5, fill = 'red')

```

Bonus: can change histogram to a probability density graph:
```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(aes(y=..density..)) + 
  geom_density()

```



5. Optional: git stage and commit

__Uses of a histogram__: Explore the distribution of a single numeric variable.


## Box plots, and violin plots

Let's make _box plots_ of population for each continent. Note: y-axis is much better on a log scale!

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ |  'x' and 'y'|
| __geometric object__  |  boxplot |
| scale                 | log y  |
| statistical transform | yes: 5-number summary |

2. Initiate the `ggplot` call, with the log y scale, and store it in the variable `a`. Print out `a`.

```{r}
#makes it easier so we don't have to rewrite all the time.
a = ggplot(gapminder, aes(continent, pop)) + 
  scale_y_log10()
a #gives blank plot. bc we still haven't added boxplot geom. 
```


3. Add the boxplot geom to `a`.

```{r}
a + geom_boxplot()
```


4. A violin plot is a kernel density on its side, made symmetric. Add that geom to `a`.
    - What's better here, boxplots or violin plots? Why?

```{r}
a + geom_violin()
#similar to kernel density

```

5. Optional: git stage and commit

__Use of boxplot__: Visualize 1-dimensional distributions (of a single numeric variable).

## Jitter plots

Let's work up to the concept of a _jitter plot_. As above, let's explore the population for each continent, but using points (again, with the y-axis on a log scale).

Let's hold off on identifying the grammar.

```{r}
a + geom_point(alpha = 0.1) 
#alpha is transparancy = 0.1 means 10 points on top of e/o to get an opaque point. This helps a bit in helping see the density of graph.
```


1. Initiate the `ggplot` call to make a scatterplot of `continent` vs `pop`; initiate the log y scale. Store the call in the variable `b`.

see above 'a'

2. Add the point geom to `b`. Why is this an ineffective plot?

see above

3. A solution is to jitter the points. Add the jitter geom. Re-run the command a few times -- does the plot change? Why?

```{r}
a + geom_jitter(alpha = 0.25)

```


4. How does the grammar differ from a box plot or violin plot?
    - ANSWER: 

5. We can add multiple geom _layers_ to our plot. Put a jitterplot overtop of the violin plot, starting with our base `b`. Try vice-versa. 

```{r}
a + geom_violin() + geom_jitter(alpha = 0.1)
#adding jitter helps you see the number of points. whereas just violin, you don't know how many data points you have. More complete picture of the data. better than pinhead plots (mean + SD) .
```


6. Optional: git stage and commit

__Uses of jitterplot__: Visualize 1-dimensional distributions, AND get a sense of the sample size.

## Time/Line Plots

Let's make some time/line plot, starting with Canada's life expectancy over time.

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ |  |
| __geometric object__  |  |
| scale                 |  |
| statistical transform |  |

skip filling in the grammar here for now...

2. In one readable call, write code that:
    1. Filters the data to Canada only
    2. Pipes the filtered data into `ggplot`
    3. Makes the time plot of `lifeExp` over time
    4. Also displays the points

```{r}
#can either pipe then store into a new variable, or just pipe it into ggplot. 
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_line() + 
  geom_point()

#geom line connects dots from left to right, even if order is different. 
#when you start using ggplot in piping, you need to use + not %>% 
```


3. Attempt to overlay line plots for all countries. That is, repeat the above code, but don't filter. What's wrong here?

```{r}
c = ggplot(gapminder, aes(year, lifeExp))
c + geom_line() #this alone doesnt give what we want. so we add the following aes:
c + geom_line(aes(group = country), alpha = 0.2)
```


4. Use the `group` aesthetic to fix the problem.

see above.

5. Optional: git stage and commit

__Uses of time/line plots__: Visualize trends of a numeric variable over time.

## Path plots

Let's see how Rwanda's life expectancy and GDP per capita have evolved over time, using a path plot.

1. Make a scatterplot. Store it in the variable `c`. 



2. We want to connect the dots from earliest point to latest. What happens if we add the "line" geom to `c`?



3. Add the appropriate geom to `c`. In that geom, specify a property of the geom: `arrow=arrow()`. 


4. Optional: git stage and commit

__Uses of path plots__: The four "corners" of the plot usually indicate different qualities. This plot allows you to see how Rwanda (or some entity) evolves over these qualities.

## Bar plots

How many countries are in each continent? Use the year 2007.

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ |  |
| __geometric object__  |  |
| scale                 |  |
| statistical transform |  |

2. After filtering the gapminder data to 2007, make a bar chart of the number of countries in each continent. Store everything except the geom in the variable `d`.



3. Notice the y-axis. Oddly, `ggplot2` doesn't make it obvious how to change to proportion. Try adding a `y` aesthetic: `y=..count../sum(..count..)`.



4. Optional: git stage, commit, and push!

__Uses of bar plots__: Get a sense of relative quantities of categories, or see the probability mass function of a categorical random variable.
