---
title: "cm011 Exercises: R as a programming language"
output: github_document
---

# Part I

## Types and Modes and Classes, Oh My!

R objects have a _type_, a _mode_, and a _class_. This can be confusing:

```{r}
a <- 3
print(typeof(a))
print(mode(a))
print(class(a))
```


```{r}
print(typeof(iris))
print(mode(iris))
print(class(iris))
```


```{r}
print(typeof(sum))
print(mode(sum))
print(class(sum))
```

Usually, there's no need to fuss about these differences: just use the `is.*()` family of functions. Give it a try:

```{r}
is.numeric(a)
is.data.frame(iris)
```

We can also coerce objects to take on a different form, typically using the `as.*()` family of functions. We can't always coerce! You'll get a sense of this over time, but try:

- Coercing a number to a character.
- Coercing a character to a number.
- Coercing a number to a data.frame. `letters` to a data.frame.

```{r}
as.character(100)
as.numeric("100")*10
as.numeric("hello") #can't convert strings to numeric. 
as.data.frame(letters)

```

There is also a slight difference between coercion and conversion, but this is usually not important.

## Vectors

Vectors store multiple entries of a data type. You'll discover that they show up just about everywhere in R, so they're fundamental and extremely important.

### Vector Construction and Basic Subsetting

We've seen vectors as columns of data frames:

```{r}
mtcars$hp #vector of horspower of cars in mtcars dataframe.

```

Use the `c()` function to make a vector consisting of the course code (`"STAT"` and `545`). Notice the coercion. Vectors must be homogeneous.

```{r}
(course = c("STAT", 545))
#second entry we input as numeric, but comes out as a character. vectors can only hold 1 type of data/info. 
```


Subset the first entry. Remove the first entry. Note the base-1 system.

```{r}
#subsetting: put the index of the entry that you want. 
#here there is a base-1 numbering system, no base 0 system here.
course[1]

course[-1] #will remove the first entry
sort(course)[1]

course #original vector remains the same. unless you reassign it...

```

Use `<-` to change the second entry to "545A". Using the same approach, add a third entry, "S01".

```{r}
#can edit entries or change entries.
course[2] = "545A" #this will change the vector entries. 
course[3] = "S01"
course[6] = "test"
course
```

Subset the first and third entry. Order matters! Subset the third and first entry.

```{r}
course[c(3,1)] #subsetting a vector with a vector. we want the 3rd entry followed by the 1st.
```

Explore integer sequences, especially negatives and directions. Especially `1:0` that might show up in loops!

```{r}
3:10 #gives a vector from 3-10.
10:-5 
1:0 #will not return a vector of length 0. 
#vector length 0: 
seq_len(0)
#vector length 10 (from 1-10)
seq_len(10)
```

Singletons are also vectors. Check using `is.vector`.

```{r}
is.vector(6)
```

```{r}
paste("hi", "you")
```


### Vectorization and Recycling

A key aspect of R is its vectorization. Let's work with the vector following vector:

```{r}
(a <- 7:-2) #reassign values in vector a. 
(n <- length(a)) #
```

Square each component:

```{r}
a^2 #squares each component of vector a. in base python we'd need to write a for loop. 
```

Multiply each component by 1 through to its length: (1st entry by 1, 2nd entry by 2, etc... )

```{r}
a*1:10 #multiply through the vector 1-10 #pairs each component, and multiplies this way.
```

It's important to know that R will silently recycle! Unless the length of one vector is not divisible by the other. Let's see:

```{r}
a*1:3 #will duplicate the 1:3 vector and keep duplicating it until it reaches the same length as a.
a*1:12 #notice how it adds two values to the vector corresponding to a[1]*11 and a[2]*12
a*1:2 #will not throw a warning, because 10/2 divides. but it still recycles over 5 times.

```


This is true of comparison operators, too. Make a vector of logicals using a comparison operator.

```{r}
a>0
```


Now try a boolean operator. Note that && and || are NOT vectorized!

```{r}
a >0 & a<5
#first does a>0 then does a<5 and then compares which one is both true, and if we have 2 true it reutrns TRUE in vector.

a >0 && a<5  #not vectorized. used in if statements because you want a true or false and not a vector. it only looks at the first entry of both vectors so a>0[1] =TRUE but a<5[1]= FALSE so overall it is false.

#use functions "all" or "any" to reduce a vector of non-1 length
#are any objects in a> 0 and <5?
any(a>0 & a<5)
all(a>0 & a<5) 
```


Recycling works with assignment, too. Replace the entire vector a with 1:2 repeated:

```{r}
a[1:n] = 1:2 #subset positions 1 to n in our vector (length 10) and we replace it with vector of length 2 which gets recycled 5 times.
a
(a[1:n]=1:2) #returns vector of length 2... we dont know why.
```


### Special Subsetting

We can subset vectors by names and logicals, too.

Recall the course vector:

```{r}
course <- c("STAT", "545A", "S01")

```

Let's give the components some names ("subject", "code", and "section") using **three** methods:

1. Using the setNames function.
- Notice that the vector does not change!!

```{r}
setNames(course,c("subject", "code", "section")) #something that sits on top of vector. can ignore this. 
```


2. Using the names function with `<-`. Also, just explore the names function.

```{r}
names(course) = c("subject", "code", "section")
course
names(course) #extracts names and returns character vector of the names if you need this.
```

3. Re-constructing the vector, specifying names within `c()`.

```{r}
course = c(subject = "STAT",code= "545",section= "S01")
course
```


Subset the entry labelled "section" and "subject". 

```{r}
#can subset vector according to names
course["section"]
course["subject"]
course[c("section", "code")]

#what happens if we have same names in vector:
foo = c(a = 5, a=6, b = 7)
foo["a"] #only returns the first value
foo[["a"]] #same as above.
```

Amazingly, we can also subset by a vector of logicals (which will be recycled!). Let's work with our integer sequence vector again:

```{r}
(a <- 7:-2)
(n <- length(a))
#subset a to all positive entries:
a[a>0] #omits all entries that don't return TRUE. 

```

## Lists

Unlike vectors, which are atomic/homogeneous, a list in R is heterogeneous. 

Try storing the course code (`"STAT"` and `545`) again, but this time in a list. Use the `list()` function.

```{r}
course = list("STAT", 545)
course #first entry a character vector of length 1 and 2nd entry a numeric vector of length 1
```


Lists can hold pretty much anything, and can also be named. Let's use the following list:

```{r}
(my_list <- list(year=2018, instructor=c("Vincenzo", "Coia"), fav_fun=typeof))
#lists can hold functions, characters, numerics, etc. 
```


Subsetting a list works similarly to vectors. Try subsetting the first element of `my_list`; try subsettig the first _component_ of the list. Notice the difference!

```{r}
my_list[1]
#subsets the list. 
my_list[[2]] #sheds that list structure.

my_list[2] #a list of length 1 that returns a vector of lenght 1.
my_list[[2]][1] #subset the first position of the vector from list position 2
```

Try also subsetting by name:

```{r}
my_list["year"]
my_list[["year"]] #returns the content of that list position. 

my_list$year #$ is common. returns whatever is in that list (ie a vector of length 1). 


```

Smells a little like `data.frame`s. It turns out a `data.frame` is a special type of list:

```{r}
(small_df <- tibble::tibble(x=1:5, y=letters[1:5]))
```

```{r}
#is it a list? YES
is.list(small_df)
```

```{r}
#let's coerce it to a list (notice the change in output in this way)
as.list(small_df)
as.vector(small_df)
as.vector(small_df)[1]
as.vector(small_df)[2]


#making a matrix:
diag(2) #in matrices you have rows and columns, but must all be the same type. 
```

Note that there's a difference between a list of one object, and that object itself! This is different from vectors.

```{r}
identical(list(4), 4) #list 4 does not = 4
identical(c(4), 4) #vector 4 = 4 (vectors dno't have extra baggage around them, lists do.)

#identical checks if two arguments are identical. 

```

# Part II

## Global Environment

When you assign variables in R, the variable name and contents are stored in an R environment called a global environment.

See what's in the Global Environment by:

- Executing `ls()`.
- Looking in RStudio, in the "Environments" pane.

```{r}

```


Making an assignment "binds" an object to a name within an environment. For example, writing `a <- 5` assigns the object `5` to the name `a` in the global environment.

The act of "searching for the right object to return" is called scoping.

By the way: the global environment is an object, too! It's the output of `globalenv()`, and is also stored in the variable `.GlobalEnv`:

```{r}
globalenv()
.GlobalEnv
```

## The Search Path

How does R know what `iris` is, yet `iris` does not appear in the global environment? What about functions like `length`, `sum`, and `print` (which are all objects, too)?

Let's explore.

1. Each package has its own environment.
    - Install and load the `pryr` package, and use `ls()` to list its bindings (its name is "package:pryr").

```{r}

```

2. There's a difference between an _environment_ and its _name_. Get the environment with name "package:pryr" using the `as.environment()` function.

```{r}

```


3. Each environment has a parent. Use `parent.env()` to find the parent of the global environment.

```{r}

```


4. There are packages that come pre-loaded with R, and they're loaded in a sequence called the search path. Use `search()` to identify that path; then see it in RStudio.

```{r}

```

First scoping rule: R looks to the parent environment if it can't find an object where it is.

5. Use `pryr::where()` to determine where the first binding to the name `iris` is located.

```{r}

```


6. Override `iris` with, say, a numeric. Now `where()` is it? Can you extract the original?

```{r}

```


7. Override `sum` with, say, a numeric. `where()` is `sum` now? Can you still use the original `sum()` function?

```{r}

```

Special scoping rule for functions! R knows whether or not to look for a function.

8. Look at the source code of the `pryr:where()` function. It contains a line that creates a binding for "env". Why, then, is `env` nowhere to be found? Answer: execution environments. 

```{r}
#pryr::where
```



