---
title: "cm008 Exercises"
output:github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse)) 
suppressPackageStartupMessages(library(gapminder))
```

## `mutate()`

create a new column from other columns. 
Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}
gapminder %>% 
  mutate(gdp = gdpPercap*pop, 
         gdpBill = round(gdp/10^9, 2)) #,2 because we want it to 2 decimal places.

```


Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
gapminder %>%
  transmute(gdp = gdpPercap*pop,
        gdpBill = round(gdp/10^9, 2)  ) #replaces all columns except the one we're adding.
```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}
if_else (letters =="h", "H", letters) 
#we took the letters vector and replaced a lowercase H with a capital letterH, and if it's already capital, then just outpute the same vector. 
gapminder %>% 
  mutate(lifeExp = if_else(country == "Canada" & year == 1952, 70, lifeExp)) %>% 
  filter (country =="Canada")
```


Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
gapminder %>%
  mutate (cc = paste(country, continent, sep = ", "))
```

These functions we've seen are called __vectorized functions__. 

## `summarize()` and `group_by()`

Use `summarize()` to compute the mean and median life expectancy using all entries:

```{r}
gapminder %>% 
  summarize(mu = mean(lifeExp),
            md = median(lifeExp))
```

Do the same thing, but try:

1. grouping by country
2. grouping by continent and country

```{r}
gapminder %>% 
  group_by(country) %>% 
  summarize(mu = mean(lifeExp),
            md = median(lifeExp))

gapminder %>% 
  group_by(continent, country) %>% 
  summarize(mu = mean(lifeExp))
#every time you use summarize the grouping gets peeled back by 1 layer??? 
#we are supposed to see a grouping info at the top of the column. 
  
```

- Notice the columns that are kept. 
- Notice the grouping listed above the tibble, especially without a call after grouping.
- Notice the peeling of groups for each summarize. 

Question: What if I wanted to keep the other numeric columns (gdpPercap, pop)? Can I? Would this even make sense?
not meaningful. 
why? we just compressed data into mean and median, so if we try to keep the old data, the population rows somehow need to be compressed into that one data point for each country. 
it wouldn't fit the dataframe. 

For each continent: What is the smallest country-wide median GDP per capita?

```{r}
gapminder %>% 
  group_by(continent, country) %>% 
  summarize(md = median(gdpPercap)) %>%  #first want median GDP per capita. but this isn't what we want. each row is its own country, but now only grouped by country because summarize peeled back the continent layer. 
  summarize(min = min(md))
  
```

Note that ggplot2's grouping is different from dplyr's! Try making a spaghetti plot of lifeExp over time for each coutry, by piping in a grouped data frame -- it won't work:

```{r}
gapminder %>% 
  group_by(country) %>% 
  ggplot(aes(year,lifeExp)) +
  geom_line()

#why doesn't this work? ggplot doesn't recognize any grouping done by dyplyr. you would need to indicate the grouping in aes. 
```


Your turn! For each continent, what is the median GDP per capita of countries with high (>60) life expectancy vs countries with low (<=60)? Sort this data frame by median GDP per capita.

```{r}
gapminder %>%
  group_by(continent, lifeExp >60) %>% #within each continent, we will have two groups (lifeExp below or greater than 60)
  summarize(md = median(gdpPercap)) %>% 
  arrange(md)
  

gapminder %>%
  mutate(age = if_else(lifeExp > 60,"high", "low" )) %>%  #if lifeExp greater than 60, then place into high or low grouping.
  group_by (continent, age) %>% #for each continent, split it up by the age group 
  summarize(md = median(gdpPercap)) %>%  #summarize function applied for median gdpPercap.
  arrange(md)
```

There are special functions to summarize by. Let's see some of them:

- `n()`: Number of rows in the group.
- `n_distinct()`

Convenience functions:

- `tally()` (= `summarize(n = n())`)
- `count(...)` (= `group_by(...) %>% tally()`)

n_distinct: How many years of record does each country have?

```{r}
gapminder %>%
  group_by(continent) %>% 
  summarize(num = n())
#count number of rows within each group using the n( ) function

gapminder %>%
  group_by(continent) %>% 
  tally()

gapminder %>%
  count(continent)

#diff ways of getting the same thing. 
```

count

## Function types

Let's stop to identify some theory of function types, and the `dplyr` approach to them.

| Function type | Explanation | Examples | In `dplyr` |
|------|-----|----|----|
| Vectorized functions | These take a vector, and operate on each component independently to return a vector of the same length. In other words, they work element-wise. | `cos`, `sin`, `log`, `exp`, `round` | `mutate` |
| Aggregate functions | These take a vector, and return a vector of length 1 | `mean`, `sd`, `length`, `typeof` | `summarize`, esp with `group_by`. |
| Window Functions | these take a vector, and return a vector of the same length that depends on the vector as a whole. | `lag`, `rank`, `cumsum` | `mutate`, esp with `group_by` |

For any generic output, we can use dplyr's `do()` function -- but that's a topic for STAT 547. 

## Grouped `mutate()`

Calculate the growth in population since the first year on record _for each country_. `first()` is useful.

```{r}
gapminder %>%
  group_by(country) %>% 
  mutate(growth = pop - first(pop)) #because the first year of growth should be 0.

lag(c(5,2,4))

gapminder %>%
  group_by(country) %>% 
  mutate(change = pop-lag(pop)) #lag takes a vector and shifts it up by 1.



```

Notice that `dplyr` has retained the original grouping.

How about growth compared to `1972`?

```{r}
gapminder %>% 
  group_by(country) %>%
  mutate(growth = pop - pop[year == 1972]) #shows the growth per year compared to year 1972.
```

Make a new variable `pop_last_time`, as the "lag-1" population -- that is, the population from the previous entry of that country. Use the `lag` function.

```{r}

```

Similar: `lead` function. The lead shifts the vector backwards 

Notice the NA's. 

## Putting it all together

Your turn: Use what we learned to answer the following questions.
[solutions: ](http://stat545.com/topics) http://stat545.com/topics

1. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent.

```{r}
gapminder %>% 
  group_by(continent, country) %>% 
  mutate (gainLifeexp = lifeExp - lag(lifeExp)) %>%  
  filter(!is.na(gainLifeexp)) %>%  #filter out the NAs
  summarize(min = min(gainLifeexp)) %>%  #if we run this and only the first line... then see the results by country. 
  summarize(min = min(min)) %>%  #summarize again to get it for the continent. 
  arrange(min)


```

2. Compute the relative gdp (NOT per capita!) of each country compared to Canada (= GDP of a country / GDP of Canada).

```{r}
cda = gapminder %>%  #select a vector of gdp per capita for canada. 
  mutate (gdp = gdpPercap * pop) %>% 
  filter (country =="Canada") %>% 
  '[['("gdpPercap") #indicate a single variable in quotation marks to extract that value. we can use this instead of select. 

 #rep will repeat that vector a certain way. so in this case it will repeat canada for each level, or value, of countries. #we assume here each country has same # of years.


Canada_compare = gapminder %>% 
  mutate(cda_gdp = rep(cda, nlevels(country)),
  gdp = gdpPercap * pop,
  rel_gdp = gdp/cda_gdp)

```

Sanity check: are Canada's numbers = 1? What is the spread of numbers like (should be small)?

```{r}
Canada_compare %>%  filter (country == "Canada")
summary(Canada_compare$rel_gdp)
```


## Summary of major one-table functions

- `select()`
- `filter()`
- `arrange()`
- `mutate()`
- `summarize()`

Together with `group_by()` and "twists" of the above. 


## Practice Exercises

Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.

1. Convert the population to a number in billions.

2. Compute the change in population from 1962 to 1972 for each country.

3. Rank the continents by GDP per capita. You should have two columns: one with continent, and another with the ranking (1 through 5). __Hint__: use the `rank()` or `min_rank()` function.
